miniKanren with fair search strategies

### 2019-04-30 Paper/paper.tex
revise

%TODO rewrite abstract
%The syntax of a programming language should reflect its semantics. When using a disjunctive operator in relational programming, a programmer would expect all sub-goals of this disjunct to share the same chance of being explored, as these clauses are written in parallel. The existing multi-arity disjunctive operator in miniKanren, however, prioritize its clauses by the order of which these clauses are written down. We have devised two new search strategies that allocate computational effort more fairly in all clauses.

@@

We propose three new search strategies, balanced interleaving DFS (biDFS), fair DFS (fDFS), and BFS. They have different characteristics of fairness (Table.~\ref{fairness}). We prove that our BFS and the BFS proposed by Seres et al~\citep{seres1999algebra} produce the same result when queried. But our code is shorter and runs faster. We also compare the efficiency of these new search strategies with the existing ones.

@@

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search strategies that is less sensitive to goal order can also be useful to little miniKanreners as well as seasoned ones. We propose two such search strategies, balanced interleaving DFS (biDFS) and breadth-first search (BFS), and observe how they affect the efficiency and the answer order of known miniKanren programs. The experiment is conducted with the miniKanren from \textit{The Reasoned Schemer, 2nd Edition}.

@@

Search strategies with almost-fair \disj{} (e.g. biDFS) give the same result in this case. However, as its name suggests, almost-fair strategies are not always fair. Fortunately, the maximal ratio of the allocated resource is bounded by a constant. Our biDFS is fair when the number of goals is a power of 2, otherwise, some goals are allocated twice more resource than the others. In following example, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So we expect lists of each sort constitute  $1/4$ of the answer list. However, search strategies with unfair \conj{} gives us much more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists of \texttt{c}) are hardly found. The situation would be worse if we add more \conde clauses. The result with biDFS, whose \conj{} is also unfair, is similar, but due to its different \disj{}, the position of \texttt{b} and \texttt{c} and swapped.

@@

A more interesting situation is when the first conjunctive goal produces infinite many states. Consider the following example, a naive specification of fair \conj{} might require search strategies to produce all sorts of singleton lists, but no longer lists, which makes the strategies \emph{incomplete}.

@@

Fairness is good but is not good enough to kick away completeness. A solution also taken in \citep{seres1999algebra} is bagging states and requiring that search spaces derived from states in the same bag are treated fairly. A natural way to bag states is by their costs. The \emph{cost} of a state is its depth in the search tree (i.e. the number of calls to relational definitions required to find them) \citep{seres1999algebra}. BFS is more than fair -- it produces answers in increasing order of cost. Running the same program gives an answer list sorted by answers' costs. In this case, the cost of an answer is equal to the length of the inner lists plus the length of the outer list.

@@

Balanced interleaving DFS (biDFS) has almost-fair \disj{} and unfair \conj{}. The implementation of biDFS differs from iDFS in the \disj{} macro. We list the new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure \texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} takes a non-empty list of goals \texttt{gs} and returns a goal. With the help of \texttt{split}, it essentially constructs a \emph{balanced} binary tree where leaves are goals in \texttt{gs} and nodes are \texttt{disj2}, whence the name of this search strategy. In contrast, the \disj{} in iDFS essentially constructs the same sort of binary tree in one of the most unbalanced forms.

@@

@@

Fair DFS (fDFS) has fair \disj{} and unfair \conj{}. The implementation of fDFS differs from iDFS's in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is changed to call a new and fair version of \texttt{append-inf}. \texttt{append-inf/fair} immediately calls its helper, \texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to \texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't been swapped. The swapping happens in the third \texttt{cond} clause of the helper, where \texttt{s?} is changed accordingly. The first two \texttt{cond} clauses essentially copy the \texttt{car}s and stop recursion when one of the input spaces is obviously finite. The third clause, as we mentioned early, is just for swapping. When the fourth and last clause runs, we know that both \texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new thunk is constructed. The new thunk calls the driver recursively. Here changing the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though it will change the order of answers). We swapped them back so that answers are produced in a more natural order.

@@

@@

Our BFS is fair in both \disj{} and \conj{}. Its implementation is based on fDFS (not iDFS). All we have to do is changing the use of \texttt{append-inf} in \texttt{append-map-inf} to \texttt{append-inf/fair}. After making this change, \texttt{append-inf} is useless.

The implementation can be improved in two aspects. First, as we mentioned in section 2.2, our BFS bag states by their cost. However, in this implementation, it is unclear where this information is recorded. Second, \texttt{append-inf/fair} is space inefficient. It makes $O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ is the ``length'' of each input search space. We address these issues in the first subsection.

Both our BFS and Seres's BFS \citep{seres1999algebra} produce answers in increasing order of cost. So it is interesting to see if they are equivalent. We prove the equivalence in Coq. The details are in the second subsection.

% The first subsection describe an optimized BFS. The optimization makes the type of search space representation more clear and reduces time-complexity of some function. The second subsection compares the pure subset of the optimized BFS with the BFS found in \citep{seres1999algebra}. We prove the two BFSs return the same answer list.

@@

As we mentioned in section 2.2, our BFS bag states by their cost. The bagging information is recorded subtly -- the \texttt{car}s of a search space have cost 0 (they are in the same bag), and the costs of states in thunk are increased by one. It is even more difficult to see \texttt{append-inf/fair} and the modified \texttt{append-map-inf} respects the cost information. We make things clear by changing the type of search space, modify related function definitions, and introducing a few more functions.

@@

Both modified functions and introduced ones are listed in Fig.~\ref{BFS-opt}. The first three functions in Fig.~\ref{BFS-opt} are search space constructor. They are newly introduced. \texttt{none} makes an empty search space. \texttt{unit} makes an space from one state. \texttt{step} makes a space from a thunk. The next function, \texttt{split}, is used to implement impure features (i.e. \texttt{ifte} and \texttt{once}). It takes a space \texttt{s-inf} and two continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some states, \texttt{ks} is called with the first state and the rest space. Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means `succeed' and `fail' respectively. The remaining functions do the same thing as before. Now it should be not difficult to see that \texttt{append-inf/fair} and \texttt{append-map-inf} do respect cost information.

@@

\begin{figure}
	\lstinputlisting{Figures/BFS-opt.rkt}	
	\caption{interface functions in optimized BFS}
	\label{BFS-opt}
\end{figure}

@@

;;TODO to be more specific?

@@

In this section, we compare the efficiency of search strategies. A concise description is in Table~\ref{compare-efficiency}. A hyphen means running out of memory. The first three benchmarks are taken from \citep{friedman_reasoned_2018}. Next two benchmarks about quine are modified from a similar test case in \citep{byrd2017unified}. The modifications are made to circumvent the need for symbolic constraint. Our version generates de Bruijnized expressions and forbids closures going into list. The two benchmarks differ in the \conde clause order of their relation interpreters. The last two benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, they differ in the \conde clause order of their relation interpreters. The first one has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the end, while the other has them at the beginning. We conjecture that iDFS would perform badly in the second case because elimination rules complicate the problem when running backward. Our statistics support our conjecture.

@@

% We devise a new search strategy, balanced interleaving DFS. The key idea is to make disjunct trees balanced. Changing the search strategy from iDFS to biDFS is not hard: 2 new functions and 1 modified macro.

% We also devise breadth-first search, whose intuition is similar to Seres's BFS. And we have proved their equivalence. We optimize our BFS with a queue.

>>>>

%TODO rewrite abstract
%The syntax of a programming language should reflect its semantics. When using 
%a disjunctive operator in relational programming, a programmer would expect 
%all 
%sub-goals of this disjunct to share the same chance of being explored, as 
%these 
%clauses are written in parallel. The existing multi-arity disjunctive operator 
%in miniKanren, however, prioritize its clauses by the order of which these 
%clauses are written down. We have devised two new search strategies that 
%allocate computational effort more fairly in all clauses.

@@

We propose three new search strategies, balanced interleaving DFS (biDFS), fair 
DFS (fDFS), and BFS. They have different characteristics of fairness 
(Table.~\ref{fairness}). We prove that our BFS and the BFS proposed by Seres et 
al~\citep{seres1999algebra} produce the same result when queried. But our code 
is shorter and runs faster. Because the two BFSs are equivalent, we do not 
distinguish them in most places, except section 6 (quantitative evaluation).

@@

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the 
%current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users 
%can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before 
%more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search 
%strategies that is less sensitive to goal order can also be useful to little 
%miniKanreners as well as seasoned ones. We propose two such search strategies, 
%balanced interleaving DFS (biDFS) and breadth-first search (BFS), and observe 
%how they affect the efficiency and the answer order of known miniKanren 
%programs. The experiment is conducted with the miniKanren from \textit{The 
%Reasoned Schemer, 2nd Edition}.

@@

Search strategies with almost-fair \disj{} (e.g. biDFS) give the same result in 
this case. However, as its name suggests, almost-fair strategies are not always 
fair. Fortunately, the maximal ratio of the allocated resource among 
disjunctive goals is bounded by a constant. Our biDFS is fair when the number 
of goals is a power of 2, otherwise, some goals are allocated twice more 
resource than the others. In following example, the clauses of \texttt{b}, 
\texttt{c}, and \texttt{d} are allocated more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute  $1/4$ of the answer list. However, 
search strategies with unfair \conj{} (e.g. iDFS, fDFS) give us much 
more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists 
of \texttt{c}) are hardly found. The situation would be worse if we add more 
\conde clauses. The result with biDFS, whose \conj{} is also unfair, is 
similar, but due to its different \disj{}, the position of \texttt{b} and 
\texttt{c} and swapped.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many states. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

@@

Fairness is good but is not good enough to kick away completeness. A solution  
is bagging states and requiring that 
search spaces derived from states in the same bag are treated fairly. A natural 
way to bag states is by their costs.
This approach is also taken in \citep{seres1999algebra}.
The \emph{cost} of a state is its depth in 
the search tree (i.e. the number of calls to relational definitions required to 
find them) \citep{seres1999algebra}. BFS is more than fair -- it produces 
answers in increasing order of cost. Running the same program gives an answer 
list sorted by answers' costs. In this case, the cost of an answer is equal to 
the length of the inner lists plus the length of the outer list.

Our definition of fair \conj{} is difference from the one in 
\citep{seres1999algebra}. They consider a \conj{} (or "selection rule" in their 
words) is fair if it ``would allow one to choose the literals in different 
order''\citep{seres1999algebra}. This requirement is not stronger than 
\emph{completeness}. We prefer our definition because it avoids overlapping 
with a known concept.

@@

@@

Balanced interleaving DFS (biDFS) has almost-fair \disj{} and unfair \conj{}. 
The implementation of biDFS differs from iDFS in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}, whence 
the name of this search strategy. In contrast, the \disj{} in iDFS essentially 
constructs the same sort of binary tree in one of the most unbalanced forms.

@@

@@

Fair DFS (fDFS) has fair \disj{} and unfair \conj{}. The implementation of fDFS 
differs from iDFS's in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, 
\texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't 
been swapped. The swapping happens at the third \texttt{cond} clause in the 
helper, where \texttt{s?} is updated accordingly. The first two \texttt{cond} 
clauses essentially copy the \texttt{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned early, is 
just for swapping. When the fourth and last clause runs, we know that both 
\texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

Our BFS is fair in both \disj{} and \conj{}. Its implementation is based on 
fDFS (not iDFS). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

The implementation can be improved in two aspects. First, as we mentioned in 
section 2.2, our BFS bag states by their cost. However, in this implementation, 
it is unclear where this information is recorded. Second, 
\texttt{append-inf/fair} is space inefficient. It makes $O(n+m)$ new 
\texttt{cons} cells every time, where $n$ and $m$ are the ``length''s of 
input search spaces. We address these issues in the first subsection.

Both our BFS and Seres's BFS \citep{seres1999algebra} produce answers in 
increasing order of cost. So it is interesting to see if they are equivalent. 
We prove the so in Coq. The details are in the second subsection.

@@

\begin{figure}
	\lstinputlisting{Figures/BFS-opt.rkt}	
	\caption{interface functions in optimized BFS (pure)}
	\label{BFS-opt}
\end{figure}

As we mentioned in section 2.2, our BFS bag states by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of states in thunk are 
computed recursively then increased by one. It is even more difficult to see 
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make things clear by changing the type of search space, modify 
related function definitions, and introducing a few more functions.

@@

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt}. The 
first three functions in Fig.~\ref{BFS-opt} are search space constructor. 
They are newly introduced. \texttt{none} makes an empty search space. 
\texttt{unit} makes an space from one state. \texttt{step} makes a space from a 
thunk. The remaining functions do the same thing as 
before. Now it should be not difficult to see that \texttt{append-inf/fair} and 
\texttt{append-map-inf} do respect cost information.

@@

\begin{figure}
	\lstinputlisting{Figures/BFS-opt-cont.rkt}	
	\caption{interface functions in optimized BFS (impure)}
	\label{BFS-opt-cont}
\end{figure}

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, is used to implement \texttt{ifte} and 
\texttt{once}. It takes a space \texttt{s-inf} and two continuations 
\texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some states, 
\texttt{ks} is called with the first state and the rest space. Otherwise, 
\texttt{kf} is called with no argument. Here `s' and `f' means `succeed' and 
`fail' respectively. The remaining functions do the same thing as before.

@@

@@

;;TODO check if reverso and appendo are absolutely the same as the ones in TRS2.

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first three benchmarks are taken from 
\citep{friedman_reasoned_2018}. Next two benchmarks about quine are modified 
from a similar test case in \citep{byrd2017unified}. The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and forbids closures going into list. The two benchmarks 
differ in the \conde clause order of their relation interpreters. The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that iDFS would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. Our statistics support our conjecture.

@@

% We devise a new search strategy, balanced interleaving DFS. The key idea is 
%to make disjunct trees balanced. Changing the search strategy from iDFS to 
%biDFS is not hard: 2 new functions and 1 modified macro.

% We also devise breadth-first search, whose intuition is similar to Seres's 
%BFS. And we have proved their equivalence. We optimize our BFS with a queue.
